---
title: "Wine Quality Prediction from Physiochemical Properties"
author: "Rebecca L.E. Miller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    keep_tex: yes
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE, echo=FALSE}
if(!require(tinytex)) install.packages("tinytex", repos = "http://cran.us.r-project.org")
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(knitr)) install.packages("knitr", repos = "https://cran.us.r-project.org")
if(!require(GGally)) install.packages("GGally", repos = "http://cran.us.r-project.org")
if(!require(corrplot)) install.packages("corrplot", repos = "http://cran.us.r-project.org")

library(knitr)
library(tinytex)
library(tidyverse)
library(caret)
library(data.table)
library(GGally)
library(corrplot)

# don't show messages for tidy summarize() calls
options(dplyr.summarise.inform = FALSE) 

# set the defaults for chunk options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE)
```

# Summary

# Introduction

The data used in this analysis is from the University of California Irvine Machine Learning Repository, data set number 1341136, or the Wine Quality data set from "Modeling Wine Preferences by Data Mining from Physiochemical Properties." \footnote{P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. Modeling wine preferences by data mining from physicochemical properties. In Decision Support Systems, Elsevier, 47(4):547-553, 2009.} The data consists of eleven physiochemical properties and a quality score between 0 and 10 for each wine in the set. There are two subsets, both for Portuguese "vinho verde" wines, one for red wine and one for white wine. The most obvious machine learning task for this dataset is to use the physiochemical data as predictors and the quality score as the response variable, or in more direct language, to use the physiochemical properties to predict quality.

## Download Data
The data set is available in two *.csv files found at [red](https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-red.csv)
and [white](https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-white.csv). The files can be downloaded with the following code.

```{r DownloadFiles, echo=TRUE, message=FALSE}
  ###-----------------------------------------------------------
  ###     Download And Parse Files From Remote Repository
  ###----------------------------------------------------------
  # create a temporary file in the local filesystem for download of raw csv
  dl <- tempfile()
  
  # download the file
  download.file("https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-red.csv", dl)
  
  # read the red wine quality file into a data frame
  red <- read.csv(file = dl, header = TRUE, sep = ";")
  
  # remove the dl file handle
  rm(dl)
  
  #--------------------- White Wine File ---------------------
  # make a new temp file
  dl1 <- tempfile()
  
  # Download raw csv
  download.file("https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-white.csv", dl1)
  
  # read the white wine quality file into a data frame
  white <- read.csv(file = dl1, header = TRUE, sep = ";")
  
  rm(dl1)
```

## Data Exploration
The physiochemical properties, or columns of the data sets, for the wines in the two files are
```{r echo=TRUE, message=FALSE }
  union(colnames(red), colnames(white))
```

Both files have the same physiochemical property columns
```{r echo=TRUE, message=FALSE}
  setdiff(colnames(red),colnames(white))
```

Neither data set contains `NA` values.
```{r}
  sum(is.na(red))
  sum(is.na(white))
```


```{r}
  corrplot(cor(red), order = "AOE", tl.col = "black")
```

# Methods

# Results

# Conclusion